Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2015-09-30 AO_bondedSi.tex
Incorporate DTJ edits to my revisions from the referee comments, up to Section 3.

Crystalline silicon is an excellent optical material for the infrared. Si has a high refractive index ($3.55-3.45$ from $\lambda = 1150-2500\;$nm at room temperature, \cite{2006SPIE.6273E..77F}) and superb transmission from slightly longer than the equivalent wavelength of its band gap ($\sim$1.15 $\mu$m) to about 6.5 $\mu$m \cite{PhysRev.108.268, PhysRev.78.178} and in the far-IR \cite{doi:10.1117/12.323764}.  It has adequate transmission  ($\alpha \sim 1$ cm$^{-1}$) for many purposes in the 18$-$30 $\mu$m region as well \cite{doi:10.1117/12.323764}.  There is an extensive industrial infrastructure for the production of ultrapure Si and for polishing and lithographic patterning and etching of optical parts.

As with a number of glasses and crystalline optical materials, it is possible to form strong physical bonds between two planar or conformal silicon surfaces.  One advantage of this direct Si$-$Si bonding of macroscopic optical parts is that it facilitates the manufacture of complex optical subsystems, for example double-convex aspheres, combinations of lenses, prisms, and transmission gratings, and pairs of gratings oriented at right angles \cite{2012SPIE.8450E..2TV, 2010SPIE.7739E.123G}.  A second important advantage is that a sufficiently close bond is optically lossless, an important consideration not only for its effect on throughput but also because close lossless bonds can eliminate concerns about optical ghosts.  The remaining advantages of bonding stem from various difficulties encountered in antireflection coating a high index material in the near and mid-infrared:  There is a limited choice of index-matching materials, in particular at longer wavelengths.  Systems often are used at cryogenic temperatures and considerations of differential thermal expansion, brittleness and hygroscopic properties can restrict the choice of coatings even further. It is hard to get good antireflection performance across broad spectral ranges and at a range of incidence angles.

There are several good reviews of Si bonding that include accounts of the history of the field and a description of process alternatives, metrology, and problems \cite{1998AnRMS..28..215G,Masteika2014}.  The foundation paper on Si bonding came in 1986 \cite{1986JAP....60.2987S}. After that, there was a sequence of papers that used IR imaging to look for evidence of defects in the bond. Stengl et al. (1988) and G{\"o}esele et al. (1995) \cite{1988JaJAP..27L2364S, 1995ApPhL..67.3614G} directly monitored the bond propagation with time-lapse IR imaging. Lehman et al. 1989 and G{\"o}esele et al. \cite{1989JaJAP..28L2141L, 1995ApPhL..67.3614G} demonstrated gap-free bonding in a micro- cleanroom, in which the bonding process occurs immediately after a rinse-spin-vacuum cycle inside of a vacuum enclosure.  Other authors investigated the growth of bubbles at bond interfaces with time and the elimination of some or all of these bubbles after annealing bonded parts at temperatures of $800-1100\;^\circ$C \cite{Horn2009, Masteika2014}.

The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy\cite{feijoo1994}, which lent credibility to the idea of distinct temperature phases and evidence for Si$-$H bonds\cite{1995ApPhA..61..101R}.  These results ultimately helped to flesh out a robust picture of the chemistry as a function of anneal temperature\cite{1996JaJAP..35.2102R, 1998AnRMS..28..215G}.  Other variables potentially affecting bond strength were studied, for example surface roughness\cite{JJAP.37.4197}, surface topography \cite{2001JOptA...3...85G}, surface preparation\cite{1996ApPhL..68.2222T}, annealing time \cite{2000JAP....88.4404H}, ambient pressure and substrate thickness \cite{1995ApPhL..67..863G, 2007ApOpt..46.6793H}, and moisture and defects \cite{2001JAP....89.6013L}.  Notably, \cite{JJAP.37.4197} found that surface roughness above 1.3 nm RMS results in poor bonding quality in their sample of wafers etched with Argon.  Greco \emph{et al.} examined the role of surface topography in determining the the bonding strength of thick optical surfaces.  These authors showed that van der Waals interactions by themselves were consistent with their available experimental evidence, but they could not rule out other forces and phenomena.  The dearth of measurements of the mean separation between contacted surfaces hindered the unambiguous identification of the forces at play in bonded, rough surfaces \cite{2001JOptA...3...85G}.

If Si$-$Si bonds are to be a useful part of the optical designer's toolbox, the bonds must have transmission losses similar to or lower than the concatenated losses of two antireflection coated surfaces over the operating wavelength (typically $< 4$\%).  For Si$-$air gaps, the full Fresnel loss is $\sim30\%$ per surface, much larger than the $4-7\;\%$ per surface of the many optical materials with refractive indices in the range $1.5-1.7$. So a 4\% transmission loss specification translates to physical gaps or bubbles having an axial extent of merely $ < 35\;$nm.  If the optical layout does not eliminate pupil ghosts from the flat Si surfaces, there are additional requirements on the reflectivity of individual bubbles in the bond. While completely bonded Si parts will be completely lossless at the interface, problems on different scales with different physical origins can lead to less-than-perfect bonds.  On the smallest scales, imperfect cleaning can leave small voids where particulates hold the bonding planes apart \cite{Mitani1990} or where hydrocarbons can serve as catalysts for gap formation.  Even in the best of cleanrooms, particles at the optically significant $10-20\;$nm scale can be present in relatively high abundance. Figure 5 of Cooper 1986 \cite{doi:10.1080/02786828608959094} shows the surface flux (particles cm$^2$ s$^{-1}$) vs. particle size ($\mu$m). There will be roughly $10^4$ times more 10 nm sized particles than 1 $\mu$m sized particles.  Another possible source of gaps at the bond is inherent roughness in the initial surfaces that leads to a failure to conform \cite{2001JOptA...3...85G}.  On larger scales, gaseous hydrogen, which is a byproduct of the hydrophilic bonding process, can migrate into the bond and form local bubbles \cite{Masteika2014}.

Most of the bonding literature concerns itself with wafer bonding, as opposed to thicker optical substrates.  In bonding wafers to each other or to thicker optically polished Si disks, one or both of the partners can conform as the bonding front propagates.  In optics, however, both of the bonding partners will likely be thicker and more mechanically rigid.  When bonding two thick substrates, small differences in the shapes of the polished surfaces can potentially lead to topological inconsistencies that prohibit some areas from bonding.  If enough pressure is placed on the substrates and if the bond strength is high enough relative to the stiffness of one or the other component, however, the surfaces will conform and bond.  Subsequent annealing will strengthen the bond.

Verification that an optical bond meets its requirements and diagnosis of possible problems in the bonding technique require an accurate assessment of the geometry and lateral and axial dimensions of any defects.  Since silicon is opaque in the optical range, the most straightforward technique for detecting defects in the bond is to image them in the infrared.  Fresnel losses will cause unbonded regions to appear darker than regions where the bond is perfect.  In cases where the axial extent of the gap is a half-wavelength or more, Newton rings appear in the image of the gap.  X-ray topography \cite{Mitani1990} exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but not of the axial extent itself. In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but also produces only limited information about the axial dimension \cite{2000RScI...71.1869G}.

We describe here a new technique for measuring the axial extent of small gaps, founded upon the ability to measure transmission as a function of wavelength with high accuracy.   A recent generation of stock near-infrared spectrophotometers like the Cary5000 from Agilent offer $0.2\;\%$ precision and good repeatability.  We use the wavelength dependence of the transmission and a model of the gap as a closely-spaced Fabry-P\'{e}rot etalon to measure axial gaps down to a few tens of nanometers in size.  In order to verify the applicability and accuracy of the technique, we have used microlithography to create small artificial gaps of known size in wafers that we then bond and measure.

For all types of gaps and bubbles, our technique needs to provide an accurate assessment of the transmission loss at the operating wavelength.  For finite-sized bubbles, measurement of the axial dimension helps us to assess the effects of high-temperature annealing \cite{Horn2009, Masteika2014} on removal of hydrogen bubbles.  A wavelength-dependent transmission variation can also indicate the presence of large numbers of gaps that are too small to be resolved spatially, as one might expect if there are many very small contaminants present along the interface.  Our technique must be able to characterize gaps with losses of a few percent at the operating wavelength.  A $2\%$ loss at $1.25\;\mu$m corresponds to a gap of $>25\;$nm at normal incidence and the gap size for a loss of this magnitude scales linearly with wavelength.  We therefore need to be able to measure transmission as a function of wavelength to a fraction of a percent, including all systematic errors, for devices operating near $1\;\mu$m and to a somewhat looser standard for devices operating at longer wavelengths.

@@

Our modeling technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si$-$vacuum interface.  The refractive index of silicon is one of the highest of any commonly used optical material\cite{2006SPIE.6273E..77F}.  We model a small Si$-$Si interface gap between two almost-bonded silicon parts as a low-Finesse Fabry-P\'{e}rot etalon \cite{2007fuph.book.....S}.  The key property of an etalon is that it has a cavity enclosed between two smooth and parallel reflective surfaces where the requirements for smoothness and parallelism depend on the reflective finesse of the cavity.  For the narrow air gaps we consider here and for the relatively modest reflectivity of a Si$-$vacuum interface, most gaps at Si$-$Si bonds will form effective etalons over most of their area.  The transmission through a Fabry-P\'{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  The reflectivity is given by Fresnel's law at normal incidence:

@@

We define the coefficient of finesse\cite{2007fuph.book.....S}, $F$ (Equation \ref{Eq:coeffF}), in the customary way to encapsulate the Fabry-P\'{e}rot etalon's dependence on reflectivity.  We calculate a coefficient of finesse in the range 2.5$-$2.6, depending on wavelength.  We assume room temperature for all calculations of refractive index.  In this work, we assume transmission is at normal incidence and the refractive index of the gap is 1.0.  Si absorbs negligibly longward of $\lambda$ = 1250 nm, which we verified experimentally by comparing the transmission of Si reference samples of different thicknesses (Figure \ref{figSiAbsorbfig}).

@@

A pair of bonded silicon wafers or pucks with a small gap forms three coupled cavities with the air gap as the the central cavity.  The axial extent of the silicon cavities are large enough---and typical measurement spectral resolution low enough---that the the Si cavities reflect and transmit light incoherently at their entrance and exit interfaces.  The multiple internal reflections within the Si therefore present a hurdle to an analysis of the air gaps.  In the air gap, however, the gap size is small enough that the cavity it forms can always be treated as a coherent structure.  We adapted coherent wave transfer techniques \cite{2007fuph.book.....S} for use with incoherent interactions \cite{2002ApOpt..41.3978K} to deal with the cavities within thick Si substrates.  The details appear in Appendix \ref{sec:Append-IMRTMM}.  We computed analytic equations for total transmission (including multiple internal reflections) for two scenarios: first a double side polished (DSP) Si sample with no gap, and second a pair of bonded Si samples with a gap thickness $d$.  The DSP Si sample with no air gap has a transmission equal to:

@@

which has an average value of about 53\%.  The top line in the upper panel of Figure \ref{figAbsoluteTrans} shows the plot of Equation \ref{eqnAbsDSPtrans} in the wavelength range $1200-2500\;$nm.

@@

where $f$ is the areal fill factor of the region exhibiting a finite gap size, and $1-f$ is therefore the areal fill factor of the perfectly bonded area.  $T_g$ is defined in Equation \ref{eqFP}.  The only free parameters in Equation \ref{eqnMix} are $d$ and $f$.  Both of these parameters are fixed in our direct bonded Si wafers with synthetic gaps, and are listed in Table \ref{tbl_meshPatterns}.  Synthetic embedded gaps therefore offer an excellent test of our ability to recover the gap axial extent $d$ and areal fill factor $f$ (Section \ref{secResults}).

@@

We evaluated our metrology method experimentally by embedding gaps of known sizes in the interface between two bonded Si substrates and measuring the effects of these gaps on IR transmission.  The substrate thicknesses ranged from 0.5 to 3.3 mm.

@@

The fill factor is the pattern area covered by gaps relative to the total pattern area.  We achieved gap sizes of $49-4000\;$nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  We did not verify the delivered fill factor, but since high precision lithography is very reliable, we assign no uncertainty to the difference between the delivered fill factor and the designed fill factor.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The stylus profilometry measured values were assigned an uncertainty of 5\%, which accorded with our experience with other etched parts.  Table \ref{tbl_experiments} lists descriptions for our bonded substrates.

@@

We cleaned the surfaces before bonding to minimize the interfacial particle density.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We also measured the large scale surface flatness of the pucks with an interferometer.  We found a typical peak to valley surface flatness of 2 waves ($\lambda=632.8\;$nm) over the central 50 mm diameter.  We prepared the substrates with standard cleaning procedures of solvents in a megasonic.  We then applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with N$_2$.  The patterned Si substrate is bonded to a blank (unpatterned) Si substrate.  We pressed the Si substrates together from the center to the outside.

>>>>

Crystalline silicon is an excellent material for infrared optics. Si has a high refractive index ($3.55-3.45$ from $\lambda = 1150-2500\;$nm at room temperature, \cite{2006SPIE.6273E..77F}) and superb transmission from slightly longer than the equivalent wavelength of its band gap ($\sim$1.15 $\mu$m) to about 6.5 $\mu$m \cite{PhysRev.108.268, PhysRev.78.178} and in the far-IR \cite{doi:10.1117/12.323764}.  It also has adequate transmission  ($\alpha \sim 1$ cm$^{-1}$) for many purposes in the 18$-$30 $\mu$m region \cite{doi:10.1117/12.323764}.  There is an extensive industrial infrastructure for the production of ultrapure Si and for polishing and lithographic patterning and etching.

As with a number of glasses and crystalline optical materials, two planar or conformal silicon surfaces can form strong physical bonds.  One advantage of this direct Si$-$Si bonding of optical parts is that it facilitates the manufacture of complex optical subsystems, for example double-convex aspheres, combinations of lenses, prisms, and transmission gratings, and pairs of gratings oriented at right angles \cite{2012SPIE.8450E..2TV, 2010SPIE.7739E.123G}.  A sufficiently close bond is optically lossless, an important consideration not only for its effect on throughput but also because close lossless bonds can eliminate concerns about optical ghosts. Other advantages of bonding stem from the difficulties encountered in antireflection coating a high index material in the near and mid-infrared:  There is a limited choice of index-matching materials, in particular at longer wavelengths.  Systems often are used at cryogenic temperatures and problems with differential thermal expansion, brittleness and hygroscopic properties can restrict the choice of coatings even further. It is hard to get good antireflection performance across broad spectral ranges and at a range of incidence angles.

There are several good reviews of Si bonding that recount the history of the field and describe process alternatives, metrology, and problems \cite{1998AnRMS..28..215G,Masteika2014}.  The foundation paper on Si bonding came in 1986 \cite{1986JAP....60.2987S}. After that, there was a sequence of papers that used IR imaging to look for evidence of defects in the bond. Stengl et al. (1988) and G{\"o}esele et al. (1995) \cite{1988JaJAP..27L2364S, 1995ApPhL..67.3614G} directly monitored the bond propagation with time-lapse IR imaging. Lehman et al. 1989 and G{\"o}esele et al. \cite{1989JaJAP..28L2141L, 1995ApPhL..67.3614G} demonstrated gap-free bonding under a vacuum in a micro- cleanroom, in which bonding occurs immediately after a rinse-spin-vacuum cycle.  Other authors investigated the growth of bubbles at bond interfaces with time and the elimination of some or all of these bubbles after annealing bonded parts at temperatures of $800-1100\;^\circ$C \cite{Horn2009, Masteika2014}.

The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy\cite{feijoo1994}, which lent credibility to the idea of distinct temperature phases and evidence for Si$-$H bonds\cite{1995ApPhA..61..101R}.  A robust picture of the chemistry as a function of anneal temperature emerged \cite{1996JaJAP..35.2102R, 1998AnRMS..28..215G}.  Other variables potentially affecting bond strength were studied, for example surface roughness\cite{JJAP.37.4197}, surface topography \cite{2001JOptA...3...85G}, surface preparation\cite{1996ApPhL..68.2222T}, annealing time \cite{2000JAP....88.4404H}, ambient pressure and substrate thickness \cite{1995ApPhL..67..863G, 2007ApOpt..46.6793H}, and moisture and defects \cite{2001JAP....89.6013L}.  Notably, \cite{JJAP.37.4197} found that surface roughness above 1.3 nm RMS results in poor bonding quality in their sample of wafers etched with Argon.  Greco \emph{et al.} examined the role of surface topography in determining the the bonding strength of thick optical surfaces.  These authors showed that van der Waals interactions by themselves were consistent with their available experimental evidence, but they could not rule out other forces and phenomena.  The dearth of measurements of the mean separation between contacted surfaces hindered the unambiguous identification of the forces at play in bonded, rough surfaces \cite{2001JOptA...3...85G}.

If Si$-$Si bonds are to be a useful part of the optical designer's toolbox, the bonds must have transmission losses similar to or lower than the concatenated losses of two antireflection coated surfaces over the operating wavelength (typically $< 4$\%).  For Si$-$air gaps, the full Fresnel loss is $\sim30\%$ per surface, much larger than the $4-7\;\%$ per surface of optical materials with refractive indices $1.5-1.7$. A 4\% transmission loss specification for Si translates to physical gaps or bubbles with an axial extent of $ < 35\;$nm.  If the optical layout does not eliminate pupil ghosts from the flat Si surfaces, there are additional requirements on the reflectivity of individual bubbles in the bond. While completely bonded Si parts will be completely lossless at the interface, problems on different scales with different physical origins can lead to less-than-perfect bonds.  On the smallest scales, imperfect cleaning can leave small voids where particulates hold the bonding planes apart \cite{Mitani1990} or where hydrocarbons can serve as catalysts for gap formation.  Even in the best of cleanrooms, particles at the optically significant $10-20\;$nm scale can be present in relatively high abundance. Figure 5 of Cooper 1986 \cite{doi:10.1080/02786828608959094} shows the surface flux (particles cm$^2$ s$^{-1}$) vs. particle size ($\mu$m). There will be roughly $10^4$ times more 10 nm sized particles than 1 $\mu$m sized particles.  Another possible source of gaps at the bond is inherent roughness in the initial surfaces that leads to a failure to conform \cite{2001JOptA...3...85G}.  On larger scales, gaseous hydrogen, which is a byproduct of the hydrophilic bonding process, can migrate into the bond and form local bubbles \cite{Masteika2014}.

Most of the bonding literature concerns itself with wafer bonding, and not thicker optical substrates.  In bonding wafers to each other or to thicker optically polished Si disks, one or both of the partners can conform as the bonding front propagates.  In optics, however, both of the bonding partners will be thicker and more rigid.  When bonding two thick substrates, small differences in the shapes of the polished surfaces could lead to topological inconsistencies that prohibit some areas from bonding.  If enough pressure is placed on the substrates and if the bond strength is high enough relative to the stiffness of one or the other component, however, the surfaces will conform and bond.  Subsequent annealing will strengthen the bond.

Verification that an optical bond meets its requirements and diagnosis of possible problems in the bonding technique require accurate measurement of the lateral and axial dimensions of any defects.  Since silicon is opaque in the optical range, the most straightforward way to detect defects in the bond is to take infrared images.  Fresnel losses will cause unbonded regions to appear darker than regions where the bond is perfect.  Newton rings appear in the image where the axial extent of the gap is a half-wavelength or more.  X-ray topography \cite{Mitani1990} exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but not of the axial extent itself. In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but also produces only limited information about the axial dimension \cite{2000RScI...71.1869G}.

We describe here a new technique for measuring the axial extent of small gaps, founded upon the ability to measure transmission as a function of wavelength with high accuracy.   A recent generation of stock near-infrared spectrophotometers like the Cary5000 from Agilent offer $0.2\;\%$ precision and good repeatability.  We use the wavelength dependence of the transmission and a model of the gap as a closely-spaced Fabry-P\'{e}rot etalon to measure axial gaps down to a few tens of nanometers in size.  In order to verify the applicability and accuracy of the technique, we used microlithography to create small artificial gaps of known size in wafers that we then bond and measure.

For all types of gaps and bubbles, our technique needs to provide an accurate assessment of the transmission loss at the operating wavelength.  For finite-sized bubbles, measurement of the axial dimension helps us to assess the efficacy of high-temperature annealing \cite{Horn2009, Masteika2014} in removal of hydrogen bubbles.  A wavelength-dependent transmission variation can also indicate the presence of large numbers of gaps that are too small to be resolved spatially, as one might expect if there are many very small contaminants present along the interface.  Our technique must be able to characterize gaps with losses of a few percent.  A $2\%$ loss at $1.25\;\mu$m corresponds to a gap of $>25\;$nm at normal incidence and the gap size for a loss of this magnitude scales linearly with wavelength.  We therefore need to be able to measure transmission as a function of wavelength to a fraction of a percent, including all systematic errors, for devices operating near $1\;\mu$m and to a somewhat looser standard for devices operating at longer wavelengths.

@@

Our modeling technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si$-$vacuum interface.  We model a small Si$-$Si interface gap between two almost-bonded parts as a low-Finesse Fabry-P\'{e}rot etalon \cite{2007fuph.book.....S}.  The key property of an etalon is that it has a cavity enclosed between two smooth and parallel reflective surfaces where the requirements for smoothness and parallelism depend on the reflective finesse.  For the narrow air gaps we consider here and for the relatively modest reflectivity of a Si$-$vacuum interface, most gaps at Si$-$Si bonds will form effective etalons over most of their area.  The transmission through a Fabry-P\'{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  The reflectivity is given by Fresnel's law at normal incidence:

@@

We define the coefficient of finesse\cite{2007fuph.book.....S}, $F$ (Equation \ref{Eq:coeffF}), in the customary way to encapsulate the Fabry-P\'{e}rot etalon's dependence on reflectivity.  $F$ has a value of 2.5$-$2.6, depending on wavelength.  We assume room temperature for all calculations of refractive index.  In this work, we assume transmission is at normal incidence and the refractive index of the gap is 1.0.  Si absorbs negligibly longward of $\lambda$ = 1250 nm, which we verified experimentally by comparing the transmission of Si reference samples of different thicknesses (Figure \ref{figSiAbsorbfig}).

@@

A pair of bonded silicon wafers or pucks with a small gap forms three coupled cavities with the air gap as the the central cavity.  The axial extent of the silicon cavities are large enough---and typical measurement spectral resolution low enough---that the the Si cavities reflect and transmit light incoherently at their entrance and exit interfaces.  In the air gap, however, the gap size is small enough that the cavity it forms can always be treated as a coherent structure.  We adapted coherent wave transfer techniques \cite{2007fuph.book.....S} for use with incoherent interactions \cite{2002ApOpt..41.3978K} to deal with the cavities within thick Si substrates.  The details appear in Appendix \ref{sec:Append-IMRTMM}.  We computed analytic equations for total transmission (including multiple internal reflections) for two scenarios: first a double side polished (DSP) Si sample with no gap, and second a pair of bonded Si samples with a gap thickness $d$.  The DSP Si sample with no air gap has a transmission equal to:

@@

which has an average value of about 53\%.  The top line in the upper panel of Figure \ref{figAbsoluteTrans} plots the value of $T_{DSP}$ from 1200 to 2500 nm.

@@

where $f$ is the areal fill factor of the region exhibiting a finite gap size, and $1-f$ is therefore the areal fill factor of the perfectly bonded area.  $T_g$ is defined in Equation \ref{eqFP}.  The only free parameters in Equation \ref{eqnMix} are $d$ (included in $T_g$) and $f$.

@@

We evaluated our metrology method experimentally by embedding gaps of known sizes in the interface between two bonded Si substrates and measuring the effects of these gaps on IR transmission.  The substrate thicknesses ranged from 0.5 to 3.3 mm.  Both $d$ and $f$ are fixed in our direct bonded Si wafers with synthetic gaps, and are listed in Table \ref{tbl_meshPatterns}.  Synthetic embedded gaps therefore offer an excellent test of our ability to recover the gap axial extent $d$ and areal fill factor $f$ (Section \ref{secResults}).

@@

The fill factor is the pattern area covered by gaps relative to the total pattern area.  We achieved gap sizes of $49-4000\;$nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  We did not verify the delivered fill factor, but since high precision lithography is very reliable, we assign no uncertainty to the difference between the delivered fill factor and the designed fill factor.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The stylus profilometry measured values were assigned an uncertainty of 5\%, based on our experience with other etched parts.  Table \ref{tbl_experiments} lists descriptions for our bonded substrates.

@@

To minimize the interfacial particle density, we cleaned the surfaces before bonding with standard cleaning procedures of solvents in a megasonic.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We also measured the large scale surface flatness of the pucks with an interferometer.  We found a typical peak to valley surface flatness of 2 waves ($\lambda=632.8\;$nm) over the central 50 mm diameter.  After cleaning, we applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with N$_2$.  The patterned Si substrate is bonded to a blank (unpatterned) Si substrate.  We pressed the Si substrates together from the center to the outside.
